Supplemental Materials for ``Universals of word order reflect optimization of grammars for efficient communication''

### 2019-10-26 writeup/pnas-word-order-si.tex
code

\paragraph{Evaluating Accuracy of Formalization}
An anonymous reviewer notes that the mapping between Dryer's relations and UD is not perfect, since some of the UD relations subsume other relations.
Here we show that this is not a problem, since the ordering of the various relations subsumed under the UD label strongly agree typologically.

TODO - say that, beyond just what Dryer reported, we see this distribution in the UD languages, and reproduce this

@@

\paragraph{Relation to Formalizations of Pragmatics}
In addition to these models, which  the concernefficiency and evolution of communication systems, there is closely related work formalizing the optimal choice of specific utterances in context.
Our work is most closely related to the Rational Speech Acts model of pragmatic reasoning \cite{frank2012predicting,goodman2013knowledge, kao2014nonliteral}.
In line with the other models discussed here, it assumes that rational speakers choose utterances to optimize informativity about the referent object, and trade this off with the cost of the utterance, which is partly chosen to be the surprisal of the utterance \cite{bennett2018extremely, hahn2018information, peloquin2019interactions}.
\citet{peloquin2019interactions} provide further discussion of the links between pragmatics and the efficiency metric (\ref{eq:eff-general}-\ref{eq:efficiency-derived}).

@@

In Study 1, we calculate parseability on the part-of-speech level, and also add part-of-speech tags when calculating predictability.
These choices are intended to prevent early overfitting during the grammar optimization process (Section~\ref{sec:neural-architectures}).
However, such unlexicalized parsers are less accurate than parsers taking acual word-forms into account, and adding part-of-speech tags might provide additional disambiguation that is absent in the original word-level input.
Here, we show that these limitations do not affect conclusions from Study 1, by replicating Study 1 with both parsers and language models operating entirely on word forms, without POS tags.
Results are shown in Figure~\ref{fig:pareto-plane-noPOS} and Table~\ref{tab:pareto-plane-noPOS}.
We compare real and baseline grammars; here, we do not have an estimate of the Pareto frontier, as the grammar optimization process uses part-of-speech tags (Section~\ref{sec:neural-architectures}).
In agreement with the previous results (Figure~\ref{fig:pareto-per-lang}), real grammars are mostly to the top right of their corresponding baselines.
We further confirm this in Figure~\ref{fig:lambda-quantile-noPOS}, which shows that most real grammars have higher efficiency than most baselines across permissible values of $\lambda$.
In fact, comparing Figure~\ref{fig:lambda-quantile-noPOS} to Figure~\ref{fig:lambda-quantile} suggests that optimality of real grammars is \emph{more} pronounced when modeling predictability and parseability fully on the level of word forms.

@@

In order to obtain baselines whose freedom is comparable to that of real languages, we constructed baselines that have the same Branching Direction Entropy \cite{futrell2015quantifying} as observed in the original corpora.
The Branching Direction Entropy measures the extent of freedom in choice between head-final and head-initial orderings, and it is a corpus-based quantitative measure of word order freedom \cite{futrell2015quantifying}.
For a given syntactic relation, the \emph{branching direction entropy} measures the entropy of the Bernoulli variable that is $1$ whenever the head is ordered before the dependent, and $0$ if the dependent is ordered before the head.
The branching direction entropy is $0$ if only one of the two orders can occur, and it is $\log 2$ if both orders are equally frequent.

@@

We compared these nondeterministic baseline grammars to their deterministic versions, for one language with relatively free order (Czech), and for two languages with relatively fixed order (English and Japanese).
Results are shown in Figure~\ref{fig:nondeterministic}.
For every one of the baseline grammars, we show both its deterministic and its nondeterministic version.
Nondeterministic grammars are less efficient than deterministic grammars, in particular in language with greater degrees of word order freedom (Czech).
This shows that deterministic grammars constitute stronger baselines.

\begin{figure}
	\centering
	\includegraphics[width=0.95\textwidth]{../results/plane/nondeterministic/pareto-plane-perLanguage-nondeterministic-mle.pdf}
	\caption{Parseability and predictability for three languages, including both deterministic (green) and nondeterministic  (blue) versions of the 50 baseline grammars.}\label{fig:nondeterministic}
\end{figure}

@@

An example is provided in Figure~\ref{fig:sent-dep}.
We show how the two grammars linearize the same syntactic dependency structure:
The correlating grammar (left) linearizes the three relation types towards the right of the head; the anti-correlating one places \emph{obj} dependencies on the left and the other dependencies on the right.
This example provides some intuitive idea of why the correlating grammar might lead to improved parseability:
Note that the red token labeled `N' occupies the same structural position in both versions.
In the anti-correlating version (right), when given only the token sequence, without the syntactic structure, this word could a priori be an \emph{obj} dependent of any of the three verbs occurring to its right.
In the correlating version (left), this `N' token can only possibly be a dependent of the verb occurring to its left.

\begin{figure}[ht]
    \centering
{
\centering{
\begin{dependency}[theme = simple]
   \begin{deptext}[column sep=1em]
	   V \& \textcolor{red}{N} \& V \& N \& V  \\
   \end{deptext}
   \depedge{1}{2}{obj}
   \depedge{2}{3}{acl}
   \depedge{3}{4}{obj}
   \depedge{1}{5}{xcomp}
\end{dependency}
\begin{dependency}[theme = simple]
   \begin{deptext}[column sep=1em]
	   \textcolor{red}{N} \& N \& V \&  V \&  V   \\
   \end{deptext}
   \depedge{4}{1}{obj}
   \depedge{1}{3}{acl}
   \depedge{3}{2}{obj}
   \depedge{4}{5}{xcomp}
\end{dependency}

>>>>

\paragraph{Evaluating Accuracy of Formalization}
An anonymous reviewer notes that the mapping between Dryer's relations and UD is not perfect, since some of the UD relations subsume other relations.
Here we provide evidence that this is not impact our conclusions, since the ordering of the various relations subsumed under the UD label strongly agree typologically.

@@

\paragraph{Relation to Formalizations of Pragmatics}
In addition to these models, which concern the efficiency and evolution of communication systems, there is closely related work formalizing the optimal choice of specific utterances in context.
Our work is most closely related to the Rational Speech Acts model of pragmatic reasoning \cite{frank2012predicting,goodman2013knowledge, kao2014nonliteral}.
In line with the other models discussed here, it assumes that rational speakers choose utterances to optimize informativity about the referent object, and trade this off with the cost of the utterance, which is partly chosen to be the surprisal of the utterance \cite{bennett2018extremely, hahn2018information, peloquin2019interactions}.
\citet{peloquin2019interactions} provide further discussion of the links between pragmatics and the efficiency metric (\ref{eq:eff-general}-\ref{eq:efficiency-derived}).

@@

In Study 1, we calculate parseability on the part-of-speech level, and also add part-of-speech tags when calculating predictability.
These choices are intended to prevent early overfitting during the grammar optimization process (Section~\ref{sec:neural-architectures}).
However, such unlexicalized parsers are less accurate than parsers taking acual word-forms into account, and adding part-of-speech tags might provide additional disambiguation that is absent in the original word-level input.
Here, we show that these limitations do not affect conclusions from Study 1, by replicating Study 1 with both parsers and language models operating entirely on word forms, without POS tags.
Results are shown in Figure~\ref{fig:pareto-plane-noPOS} and Table~\ref{tab:pareto-plane-noPOS}.
We compare real and baseline grammars; here, we do not have an estimate of the Pareto frontier, as the grammar optimization process uses part-of-speech tags (Section~\ref{sec:neural-architectures}).
In agreement with the previous results (Figure~\ref{fig:pareto-per-lang}), real grammars are mostly to the top right of their corresponding baselines.
We further confirm this in Figure~\ref{fig:lambda-quantile-noPOS}, which shows that most real grammars have higher efficiency than most baselines across permissible values of $\lambda$.
In fact, comparing Figure~\ref{fig:lambda-quantile-noPOS} to Figure~\ref{fig:lambda-quantile} suggests that optimality of real grammars is \emph{more} pronounced when modeling predictability and parseability fully on the level of word forms.
%This might be because such word-level models can account for morphology.

@@

In order to obtain baselines whose freedom is comparable to that of real languages, we constructed baselines that have the same Branching Direction Entropy \cite{futrell2015quantifying} as observed in the original corpora.
The Branching Direction Entropy measures the extent of freedom in choice between head-final and head-initial orderings, and it is a corpus-based quantitative measure of word order freedom \cite{futrell2015quantifying}.
For a given syntactic relation, its branching direction entropy measures the entropy of the Bernoulli random variable that is $1$ whenever the head is ordered before the dependent, and $0$ if the dependent is ordered before the head.
The branching direction entropy is $0$ if only one of the two orders can occur, and it is $\log 2$ if both orders are equally frequent.

@@

We compared these nondeterministic baseline grammars to their deterministic versions, for one language with relatively free order (Czech), and for two languages with relatively fixed order (English and Japanese).
Results are shown in Figure~\ref{fig:nondeterministic}.
For every one of the baseline grammars, we show both its deterministic and its nondeterministic version.
Nondeterministic grammars are less efficient than deterministic grammars, in particular in languages with greater degrees of word order freedom (Czech).
This shows that deterministic grammars constitute stronger baselines.

\begin{figure}
	\centering
	\includegraphics[width=0.95\textwidth]{../results/plane/nondeterministic/pareto-plane-perLanguage-nondeterministic-mle.pdf}
	\caption{Parseability and predictability for three languages, including both deterministic (green, light) and nondeterministic  (blue, dark) versions of the 50 baseline grammars.}\label{fig:nondeterministic}
\end{figure}

@@

An example is provided in Figure~\ref{fig:sent-dep}.
We show how the two grammars linearize the same syntactic dependency structure:
The correlating grammar (left) linearizes the three relation types towards the right of the head; the anti-correlating one places \emph{obj} dependencies on the left and the other dependencies on the right.
This example provides some intuitive idea of why the correlating grammar might lead to improved parseability:
Note that the red boldface token labeled `N' occupies the same structural position in both versions.
In the anti-correlating version (right), when given only the token sequence, without the syntactic structure, this word could a priori be an \emph{obj} dependent of any of the three verbs occurring to its right.
In the correlating version (left), this `N' token can only possibly be a dependent of the verb occurring to its left.

\begin{figure}[ht]
    \centering
{
\centering{
\begin{dependency}[theme = simple]
   \begin{deptext}[column sep=1em]
	   V \& \textbf{\textcolor{red}{N}} \& V \& N \& V  \\
   \end{deptext}
   \depedge{1}{2}{obj}
   \depedge{2}{3}{acl}
   \depedge{3}{4}{obj}
   \depedge{1}{5}{xcomp}
\end{dependency}
\begin{dependency}[theme = simple]
   \begin{deptext}[column sep=1em]
	  \textbf{\textcolor{red}{N}} \& N \& V \&  V \&  V   \\
   \end{deptext}
   \depedge{4}{1}{obj}
   \depedge{1}{3}{acl}
   \depedge{3}{2}{obj}
   \depedge{4}{5}{xcomp}
\end{dependency}
